Bayesian model comparison for rare variant association studies of multiple phenotypes

### 2018-01-27 tex/manuscript.tex
Text updates

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Whole genome sequencing studies applied to large populations or biobanks with extensive phenotyping raise new analytic challenges. The need to consider many variants at a locus or group of genes simultaneously and the potential to study many correlated phenotypes with shared genetic architecture provide opportunities for discovery and inference that are not addressed by the traditional one variant-one phenotype association study. Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies. In so doing we consider the use of summary statistic data to apply 1) standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, which may be used in practice for downstream quality control, and 3) models for identifying protective protein-truncating variants, which can expedite drug discovery.  Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals. Finally, we demonstrate its use with rare variant data combined with asthma diagnosis and haematological and spirometry measures for individuals in the UK Biobank. We show that we are able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.

@@

%% Genome variant annotation and protein truncating variants and  protective modifiers of disease risk
Aggregation techniques rely on variant annotations to assign variants to groups for analysis. MRP allows for the inclusion of priors on the spread of effect sizes that can be adjusted depending on what type of variants are included in the analysis. For instance, protein truncating variants (PTVs)\cite{rivas2013assessing,rivas2015effect} are an important class of variants that are more likely to be functional because they often disrupt the normal function of a gene. This biological knowledge can be reflected in the choices of priors for PTVs in MRP. Since PTVs typically abolish gene function, there is particular interest in identifying protective PTV modifiers of human disease risk that may serve as targets for therapeutics\cite{pcsk9,cohen2006sequence,sullivan2012effect}. We therefore demonstrate how the MRP model comparison approach can improve discovery of such protective signals by modeling the location (direction) of genetic effects which prioritizes variants or genes that are consistent with protecting against disease.

@@

%%Apply methods to the UK Biobank
We also apply MRP to summary statistics for two groups of related phenotypes from the UK Biobank. First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), Forced Expiratory Volume in 1-second (FEV$_1$, INI3063), and Forced Vital Capacity (FVC, INI3062). We find that XXX. We next apply MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find strong evidence that rare PTVs in \textit{ANGPTL7} protect against glaucoma. These analyses show that MRP recovers results from typical single variant-single phenotype analyses while identifying new rare variant associations that include protective modifiers of disease risk.

@@

\subsection*{UK Biobank Data}
\subsubsection*{GWAS Summary Statistics}
We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as previously described~\cite{DeBoever179762}. For asthma, we used the Firth-fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in (Hill 2017), but switches to a port of logistf() (https://cran.r-project.org/web/packages/logistf/index.html) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps. We used the following covariates in our analysis: age, sex, array type, and the first four principal components, where array type is a binary variable that represents whether an individual was genotyped with UK Biobank Axiom Array or UK BiLEVE Axiom Array. For variants that were specific to one array, we did not use array as a covariate. Asthma cases were defined using both Hospital Episode Statistics and verbal questionnaire responses. We used the provided values from the UK Biobank for eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV$_1$). The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, Forced Expiratory Volume in 1-second (FEV$_1$)=INI3063, and Forced Vital Capacity (FVC)=INI3062) correspond to the phenotype codes used in on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].

@@

\subsubsection*{Model comparison applied to glaucoma and related phenotypes}
We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing joint analyses. Elevated intra-ocular pressure is a measure of the fluid pressure in the eye. Elevated intra-ocular pressure is associated with glaucoma risk and has been linked to genetic variants associated with glaucoma \cite{doi:10.1001/jama.2014.3192}. Corneal resistance factor is a measure of the cornea's ability to resist mechanical stress and has been associated with glaucoma presence and severity \cite{CXO:CXO414, MANSOURI2012419, GriseDulac:2012bo}. While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7}, \textit{WNT10A}, and \textit{KLHL22} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-E). Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} but not \textit{KLHL22}
These results illustrate the ability of applying model comparison to multiple traits simultaneously to identify protective modifiers of disease.

@@

\section*{Discussion}
In this study, we developed a Bayesian model comparison approach MRP that shares information across both variants and phenotypes to identify rare variant associations. We used simulations to compare MRP to the widely used burden and SKAT tests for identifying rare variant associations and found that jointly considering both variants and phenotypes can improve the ability to detect associations. We also applied the MRP model comparison framework to summary statistic data from four traits from the UK Biobank: asthma, eosinophil counts, forced expiratory volume in 1-second (FEV$_1$), and forced vital capacity (FVC). We identified strong evidence for the previously described association between \textit{IL33} and asthma\cite{DeBoever179762, 10.1371/journal.pgen.1006659} as well as evidence of association for several additional genes that have not been previously reported (Table~\ref{table2}).

@@

%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
\paragraph*{S3 Fig.}
\label{asthma_similar}
{\bf Results for independent effects model applied to asthma, eosinophil counts, FEV$_1$, and FVC.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects. The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot. Only $\log_{10}$ Bayes Factors greater than -5 are plotted.

@@

% I added in the Author Contributions section
\section*{Author Contributions}
M.A.R. and M.P. designed the method and derived all analytical calculations. M.A.R., M.P., and C.D. wrote the manuscript. Y.T., M.A., and C.D. provided analysis and designed figures. T.P. designed hDF5 tables and implementation of loaders. M.J.D. provided critical feedback on methodology. TODO: Add C. Spencer contributions.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Whole genome sequencing studies applied to large populations or biobanks with extensive phenotyping raise new analytic challenges. The need to consider many variants at a locus or group of genes simultaneously and the potential to study many correlated phenotypes with shared genetic architecture provide opportunities for discovery and inference that are not addressed by the traditional one variant-one phenotype association study. Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies. We consider the use of summary statistic data to apply 1) standard univariate and multivariate gene-based meta-analysis models, 2) models for assessing heterogeneity of genetic effects, and 3) models for identifying protective protein-truncating variants that can expedite drug discovery.  Through simulation studies, we demonstrate that the proposed model comparison approach may improve ability to detect rare variant association signals. Finally, we demonstrate the model on two groups of phenotypes from the UK Biobank: 1) asthma diagnosis, eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor. We demonstrate that the model leverages information from multiple phenotypes to identify known and novel rare variant associations for asthma and glaucoma. Overall the model comparison approach is able to retain useful features from widely-used meta-analysis approaches and prioritize protective modifiers of disease risk.

@@

%% Genome variant annotation and protein truncating variants and  protective modifiers of disease risk
Aggregation techniques rely on variant annotations to assign variants to groups for analysis. MRP allows for the inclusion of priors on the spread of effect sizes that can be adjusted depending on what type of variants are included in the analysis. For instance, protein truncating variants (PTVs)\cite{rivas2013assessing,rivas2015effect} are an important class of variants that are more likely to be functional because they often disrupt the normal function of a gene. This biological knowledge can be reflected in the choices of priors for PTVs in MRP. Since PTVs typically abolish or severely alter gene function, there is particular interest in identifying protective PTV modifiers of human disease risk that may serve as targets for therapeutics\cite{pcsk9,cohen2006sequence,sullivan2012effect}. We therefore demonstrate how the MRP model comparison approach can improve discovery of such protective signals by modeling the location (direction) of genetic effects which prioritizes variants or genes that are consistent with protecting against disease.

@@

%%Apply methods to the UK Biobank
We applied MRP to summary statistics for two groups of related phenotypes from the UK Biobank. First, we apply MRP to asthma (HC382: the corresponding phenotype label in Global Biobank Engine [\url{https://biobankengine.stanford.edu}]), eosinophil count (INI30150), forced expiratory volume in 1-second (FEV$_1$, INI3063), and forced vital capacity (FVC, INI3062). We recover the reported association between a rare PTV in \textit{IL33} and asthma as well as a new association between \textit{CCR3} and asthma \cite{DeBoever179762, 10.1371/journal.pgen.1006659}. We also applied MRP to glaucoma (HC276), intra-ocular pressure (INI5263), and corneal resistance factor (INI5265) and find evidence that rare coding variants in \textit{ANGPTL7} and \textit{KLHL22} may protect against glaucoma. These analyses show that MRP recovers results from typical single variant-single phenotype analyses while identifying new rare variant associations that include protective modifiers of disease risk.

@@

\subsection*{UK Biobank Data}
\subsubsection*{GWAS Summary Statistics}
We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as previously described \cite{DeBoever179762}. For asthma, we used the Firth fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in (Hill 2017), but switches to a port of logistf() (https://cran.r-project.org/web/packages/logistf/index.html) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps. We used the following covariates in our analysis: age, sex, array type, and the first four principal components, where array type is a binary variable that represents whether an individual was genotyped with UK Biobank Axiom Array or UK BiLEVE Axiom Array. For variants that were specific to one array, we did not use array as a covariate. Asthma cases were defined using both Hospital Episode Statistics and verbal questionnaire responses. We used the provided values from the UK Biobank for eosinophil counts, forced vital capacity (FVC), and forced expiratory volume in 1-second (FEV$_1$). The phenotype codes used throughout (asthma=HC382, eosinophil count=INI30150, Forced Expiratory Volume in 1-second (FEV$_1$)=INI3063, and Forced Vital Capacity (FVC)=INI3062) correspond to the phenotype codes used in on the Global Biobank Engine [\url{https://biobankengine.stanford.edu}].

@@

\subsubsection*{Model comparison applied to glaucoma and related phenotypes}
We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing joint analyses. Elevated intra-ocular pressure is a measure of the fluid pressure in the eye. Elevated intra-ocular pressure is associated with glaucoma risk and has been linked to genetic variants associated with glaucoma \cite{doi:10.1001/jama.2014.3192}. Corneal resistance factor is a measure of the cornea's ability to resist mechanical stress and has been associated with glaucoma presence and severity \cite{CXO:CXO414, MANSOURI2012419, GriseDulac:2012bo}. While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7}, \textit{WNT10A}, and \textit{KLHL22} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-E). Applying the MRP model to identify protective effects also identified \textit{ANGPTL7} and \textit{WNT10A} but not \textit{KLHL22}. Expression of \textit{ANGPTL7} is upregulated in glaucome and has been proposed to regulate intra-ocular pressure and glaucoma risk \cite{Comes:2011ex, Kuchtey:2008ek}. \textit{KLHL22} has not been previously associated with glaucoma though a suggestive association was reported for retinopathy in individuals without diabetes \cite{10.1371/journal.pone.0054232}. These results illustrate that applying model comparison to multiple traits simultaneously can prioritize the discovery of protective modifiers of disease.

@@

\section*{Discussion}
In this study, we developed a Bayesian model comparison approach MRP that shares information across both variants and phenotypes to identify rare variant associations. We used simulations to compare MRP to the widely used burden and SKAT tests for identifying rare variant associations and found that jointly considering both variants and phenotypes can improve the ability to detect associations. We also applied the MRP model comparison framework to summary statistic data from two groups of traits from the UK Biobank: asthma diagnosis, eosinophil counts, FEV$_1$, and FVC; and glaucoma diagnosis, intra-ocular pressure, and corneal resistance factor. We identified strong evidence for the previously described association between the PTV rs146597587 in \textit{IL33} and asthma\cite{DeBoever179762,10.1371/journal.pgen.1006659} as well as evidence for a potentially novel association for \textit{CCR3} (Table~\ref{asthma_table}). We also identified genetic evidence for a link between rare variants in \textit{ANGPTL7} and glaucoma, consistent with previous experiments that suggested a role for \textit{ANGPTL7} in glaucoma \cite{Comes:2011ex, Kuchtey:2008ek}. These results demonstrate the ability of the MRP model comparison approach to leverage information across multiple phenotypes and variants to discover rare variant associations.

@@

%% FIGURE: ASTHMA MANHATTAN AND FOREST FOR INDEPENDENT EFFECTS
\paragraph*{S3 Fig.}
\label{asthma_similar}
{\bf Results for similar effects model applied to asthma, eosinophil counts, FEV$_1$, and FVC.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) asthma (HC382), (B) eosinophil counts (INI30150), (C) forced vital capacity (FVC, INI3062), (D) forced expiratory volume in 1-second (FEV$_1$, INI3063), (E) all four traits jointly, and (F) all four traits jointly with focus on protective effects. The four genes outside of chromosome 6 with the largest Bayes Factors greater than three are labeled in each plot. Only $\log_{10}$ Bayes Factors greater than -5 are plotted.

\paragraph*{S4 Fig.}
\label{glaucoma_similar}
{\bf Results for similar effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP similar effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly. (E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma. The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows. Only $\log_{10}$ Bayes Factors greater than zero are plotted.

\paragraph*{S5 Fig.}
\label{glaucoma_independent}
{\bf Results for independent effects model applied to glaucoma intra-ocular pressure, and corneal resistance factor.} $\log_{10}$ Bayes Factors from applying MRP independent effects model to summary statistics for missense and protein-truncating variants from (A) glaucoma (HC276), (B) intra-ocular pressure (INI5263), (C) corneal resistance factor (INI5265), and (D) all three traits jointly. (E) shows the results of a joint analysis focused on finding rare variants that protect against glaucoma. The genes outside of chromosome 6 with with Bayes Factor greater than three are indicated by arrows. Only $\log_{10}$ Bayes Factors greater than zero are plotted.

@@

% I added in the Author Contributions section
\section*{Author Contributions}
MAR and MP designed the method and derived all analytical calculations. MAR, MP, and CD wrote the manuscript. CCAS, YT, MA, and CD provided analysis and designed figures. TP designed HDF5 tables and implementation of loaders. MJD provided critical feedback on methodology.
